---
title: "Tailor package"
author: "Matei Ionita"
date: "6/21/2020"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tailor package}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Introduction

In recent years there have been many proposals for clustering approaches in the study of flow cytometry data. 
Tailor implements a clustering workflow based on Gaussian mixture modeling, motivated by the ability of this 
approach to identify distinct cell populations which show up as the tail of a unimodal distribution. 
In contrast, many existing clustering approaches would consider such tails as noise, and consider the 
entire unimodal distribution as a single population. 

See, for example, the case of PD1 in the
example dataset. The leftmost panel shows a kernel density estimate of the global PD1 distribution,
exhibiting the tail behavior described above. The rightmost panel shows how Tailor models this data
as a mixture of two Gaussian components. The middle panel shows the mixed distribution, which closely
approximates the original density estimate.

```{r fig_tail, fig.width=8, fig.height=4, echo=FALSE}
library(Tailor)
fileName <- system.file("extdata", "sampled_flowset_old.rda", package = "Tailor")
load(fileName)   # fs_old
data <- fs_old[[4]]

param <- flowCore::colnames(fs_old)[19]
tailor_params <- flowCore::colnames(fs_old)[c(19, 20)]
mixtures_1D <- get_1D_mixtures(data, tailor_params, max_mixture = 2)

inspect_1D_mixtures(data, mixtures_1D, param)
```

In general, Gaussian mixture modeling has two disadvantages: slow runtime, and sensitivity to initialization. 
Tailor mitigates both of these with the help of a preliminary binning step before the main mixture modeling 
algorithm (see section on Advanced workflow for details). The main algorithm is run on a sample of 
representatives from all bins; the choice of these representatives should be thought of as a 
density-dependent subsampling of the data. Moreover, the mixture components are initialized with mean 
equal to the mean of the largest bins, ensuring an initialization which captures most of the 
biologically interesting variation within the data.


## Summary
The Tailor package contains methods that accomplish three purposes:

* learning a model of the data: `tailor_learn()`, `get_1D_mixtures()`, `customize_1D_mixtures()`, `categorical_labeling()`

* predicting the most likely cluster membership for a batch of data: `tailor_predict()`

* visualization: `plot_tailor_majpheno()`, `plot_tailor_fluorescence()`, `inspect_1D_mixtures()`


## Default workflow
Some preparations: load the package, some data, choose the parameters to be used in the 
clustering analysis, and choose a seed for the random generator. For a quick peek at the dataset, 
plot a kernel density estimate for each parameter.

```{r preparations, fig.align = "center", fig.width=8, fig.height=5}
library(Tailor)
fileName <- system.file("extdata", "sampled_flowset_old.rda", package = "Tailor")
load(fileName)   # fs_old

tailor_params <- flowCore::colnames(fs_old)[c(7:9, 11:22)]
data <- fs_old[[4]]
set.seed(524)

plot_kdes_global(data, params = tailor_params)
```


The simplest way to learn a Tailor model is to feed the data directly to the `tailor_learn()` 
method. (See the Advanced workflow section below for more complex options.)

```{r tailor_vanilla}
n_components <- 50

tailor_obj <- tailor_learn(data = data,
                          params = tailor_params,
                          mixture_components = n_components)
```

Tailor attempts to model the data with the user-specified number of mixture components. 
The number of clusters you eventually obtain is likely to be smaller, for two reasons:

* Some mixture components are dropped at initialization stage, if they contain too few events. 
The user is warned when this happens.

* Some mixture components are merged into the same categorical cluster; more details below.

Because of this, we recommend overestimating, rather than underestimating `mixture_components`.

The `mixture` field of the Tailor object contains information about the 50 mixture components 
used to model the data.
`mixture$pro`, `mixture$means` and `mixture$variance$sigma` contain the mixture proportions, 
means and variances, respectively.
For example, below we access the mixture proportion and mean of the first mixture component. 
We can see that this is a large component, containing around 35% of the data, and consists of 
helper T cells, as shown by the high CD3 and CD4 expression.

```{r tailor_mixture}
tailor_obj$mixture$pro[1]
tailor_obj$mixture$mean[1,]
```

Tailor clusters, also called categorical clusters, are unions of two or more mixture components 
which have similar expression for all parameters. Information about categorical clusters can 
be found in the `cat_clusters` field of the Tailor object.

```{r tailor_categorical}
head(tailor_obj$cat_clusters$phenotypes)
head(tailor_obj$cat_clusters$centers)

# a list mapping each of the 50 mixture components to a categorical cluster
tailor_obj$cat_clusters$mixture_to_cluster
```

If available, load a file containing definitions of major phenotypes (e.g. Naive CD4 T cell, Macrophage). 
You need to provide a relative measure of fluorescence intensity: "lo", "hi", "dc"(= "don't care"), 
and Tailor chooses thresholds adaptively based on the data.

We can use this to enhance the Tailor object, labeling each categorical cluster by a major phenotype.

```{r categorical_labeling}
defsfile <- system.file("extdata", "pheno_definitions.csv", package = "Tailor")
defs <- read.csv(defsfile, row.names = 1, stringsAsFactors=FALSE)
defs

tailor_obj <- categorical_labeling(tailor_obj, defs)
```

Given the Tailor object computed with `tailor_learn()`, we can use `tailor_predict()` to assign 
cluster membership to data. In this example, we use the same data that was used in the learning step. 
But `tailor_predict()` can be used just as well with new data.

```{r tailor_predict}
tailor_pred <- tailor_predict(data = data,
                             tailor_obj = tailor_obj)
```

`tailor_predict()` returns two lists. `$mixture_mapping` assigns each datapoint to a mixture component. `$cluster_mapping` assigns each datapoint to a categorical cluster.

Use `plot_tailor_majpheno()` to view a t-SNE embedding to 2D of the categorical cluster centroids. 
They are color-coded by major phenotype, and the size of the dots depends logarithmically on cluster size.

```{r tailor_plot_clusters, fig.align= "center", fig.width=6, fig.height=4}
set.seed(524)
plot_tailor_majpheno(tailor_obj)
```

Alternatively, plot the same t-SNE embeddings, color-coded by (bi-exponentially transformed) mean 
fluorescence intensity of the cluster. 

```{r tailor_plot_mfi, fig.align= "center", fig.width=9, fig.height=9}
set.seed(524) # use the same seed as above, so figures can be compared
plot_tailor_fluorescence(tailor_obj)
```

## Advanced workflow

You will get better results if you take control over the initial step (or "binning step") of 
Tailor's learning phase.
In this step, events are placed into phenotypically similar bins, and one or more representatives 
from each bin is chosen. You should think of binning as a way to perform density-based subsampling 
of the data. To define the bins, Tailor models the distribution of each individual parameter as a 
Gaussian mixture, with 1-3 components. (This is distinct from the main modeling step, where the 
15-dimensional distribution of all parameters is modeled as a Gaussian mixture with 50 components.) 
This 1D modeling happens behind the scenes in a run of `tailor_learn()` with default settings. 
Instead, you can do it explicitly using `get_1D_mixtures()`, and visualize the result with 
`inspect_1D_mixtures()`.

```{r 1D_mixtures, fig.width=8, fig.height=4}
mixtures_1D <- get_1D_mixtures(data, tailor_params,
                              max_mixture = 3)

inspect_1D_mixtures(data, mixtures_1D, tailor_params)
```

Most of the 1D models look reasonable. But, given some domain knowledge, it may seem more reasonable to
model CD27 as one mixture component instead of two, and CD14 as two mixture components instead of 
three. We can use the method `customize_1D_mixtures()` to update a user-specified 
list of parameters. Compare the result to the previous models for CD27 and CD11 above.

```{r customize_1D_mixtures, fig.width=8, fig.height=4}
to_customize <- list("CD27AF700" = 1, "CD11BAPCCY7" = 2)
mixtures_1D <- customize_1D_mixtures(data, to_customize, mixtures_1D, verbose = TRUE)
inspect_1D_mixtures(data, mixtures_1D, names(to_customize))
```

We can pass the 1D mixtures created above as an argument to `tailor_learn()`, and they will be used 
in the binning process. We can also specify some 1D mixture components that should be merged in the 
same bin, if we have grounds to believe that they represent the same biological population.

```{r learn_advanced}
# mixtures_1D$to_merge <- list("CD11BAPCCY7"= c(1,2), "CD45RAQ655" = c(1,2), "CD27AF700" = c(2,3))

tailor_obj <- tailor_learn(data = data,
                          params = tailor_params,
                          mixtures_1D = mixtures_1D,
                          mixture_components = n_components)

tailor_obj <- categorical_labeling(tailor_obj, defs)
tailor_pred <- tailor_predict(data = data,
                             tailor_obj = tailor_obj)
```

We can visualize the t-SNE embeddings of the new categorical clusters.

```{r tailor_plot_clusters_advanced, fig.align= "center", fig.width=6, fig.height=4}
set.seed(524)
plot_tailor_majpheno(tailor_obj)
```

```{r tailor_plot_fluorescence_advanced, fig.align="center", fig.width=9, fig.height=9}
set.seed(524)
plot_tailor_fluorescence(tailor_obj)
```


## Remarks

The Tailor package is a work in progress. When you find something that doesn't work, please 
let me know.

